Verified Nanopasses for Compiling Conditionals

### 2025-06-20 paper.tex
tweaks

\usepackage{mathpartir}
\usepackage{listings}
\usepackage{stmaryrd}

@@

The compilation of Boolean expressions and conditionals is a classic
topic in the compiler literature, going back to an algorithm of
\citet{Aho:1986qf} (presented in the form of an attribute grammar)
that translates Boolean expressions and conditionals into gotos. The
compilation of Boolean expressions and conditionals often occurs at an
interesting stage in the compilation pipeline, where the intermediate
language changes from an abstract syntax \emph{tree} into a control
flow \emph{graph}, or equivalently, an intermediate language with
\texttt{goto}. Further, when targeting an assembly language such as
Intel's x86, the naive approach to compiling boolean expressions and
conditionals produces embarassingly inefficient code. Similar issues
arise in the compilation of languages such as Haskell, where the
case-of-case transformation is employed to optimize dispatching on
algebraic datatypes \citep{PEYTONJONES19983}.

Olivier Danvy, in his paper \emph{A New One-Pass Transformation into
Monadic Normal Form}~\citep{Danvy:2003fk}, presents a beautiful family
of structurally recursive functions that translate Boolean expressions
and conditionals into a monadic normal form that (1) no longer
contains Boolean expressions, (2) the condition expression of a
condition is a value, and (3) jumps are expressed using zero-arity
function calls. One of the key parts of that translation is in the
compilation of a conditional expression such as $(\IF{b}{e_1}{e_2})$,
where the expression $b$ is compiled using a special function
$\mathcal{B}$ that takes two more arguments, the generated code for
the two branches, that is, for the two continuations of $b$. Thus,
$\mathcal{B}$ can analyze the expression $b$ to decide how to generate
code for the $\mathsf{if}$, or whether an $\mathsf{if}$ is even
necessary.
\[
\CE{\IF{b}{e_1}{e_2}} = \CB{b}{\CE{e_1}}{\CE{e_2}}
\]
For example, in the simple cases where $b$ is $\TRUE$ or $\FALSE$,
no branching is needed.
\begin{align*}
\CB{\TRUE}{c_1}{c_2} &= c_1 \\
\CB{\FALSE}{c_1}{c_2} &= c_2
\end{align*}

@@

\subsection{Structured Imperative IR}

@@

data CProg : Set where
  Program : $ℕ$ $→$ Id $→$ List CStmt $→$ CProg
\end{lstlisting}

Note that the \lstinline{CStmt} type represents not just one statement
but a block of statements that always end in either a
\lstinline{Return} statement (that halts the program) or a jump to
another block (\lstinline{Goto} or \lstinline{IfEq}).

We choose a big-step semantics for C-like language because the
semantics is no longer structurally recursive and furthermore,
programs are no longer guaranteed to terminate.\footnote{The programs
in the image of Explicate Control are guaranteed to terminate, but
Agda doesn't know that.}  So encoding the semantics as a function in
Agda would come with complications (such as using a gas parameter)
that can be side-stepped by using a big-step semantics. Compared to
the semantics for the imperative IR, this semantics refers to a list
of blocks $B$ that can be the targets of gotos.

\begin{lstlisting}
data _,_$⊢ᶜ$_$⇓$_$⊣$_ : StateImp $→$ List CStmt $→$ CStmt
                      $→$ Value $→$ StateImp $→$ Set
\end{lstlisting}

The rule for \lstinline{Goto} looks up the block for label $l$ and
executes that block.
\[
\inferrule{s , B ⊢ᶜ B[l] ⇓ v ⊣ s′}
          {s , B ⊢ᶜ \mathtt{Goto}~ l ⇓ v ⊣ s′}
\]
\noindent The rule for \lstinline{IfEq}, in the case when the
arguments are equal, is as follows:
\[
\inferrule{\texttt{interp-CExp}(\mathtt{Eq}~a_1~a_2) = \mathsf{just}~\mathsf{true} \\
           s , B ⊢ᶜ B[l_1] ⇓ v ⊣ s′}
          {s , B ⊢ᶜ \mathtt{IfEq}~a_1~a_2~l_1~l_2 ⇓ v ⊣ s′}
\]

@@

@@

@@

>>>>

\usepackage{url}
\usepackage{mathpartir}
\usepackage{listings}
\usepackage{stmaryrd}

@@

The compilation of Boolean expressions and conditionals is a classic
topic in the compiler literature, going back to an algorithm of
\citet{Aho:1986qf} (presented in the form of an attribute grammar)
that translates Boolean expressions and conditionals into gotos. The
compilation of Boolean expressions and conditionals often occurs at an
interesting stage in the compilation pipeline, where the intermediate
language changes from an abstract syntax \emph{tree} into a control
flow \emph{graph}, or equivalently, an intermediate language with
\texttt{goto}. Further, when targeting an assembly language such as
Intel's x86, the naive approach to compiling boolean expressions and
conditionals produces embarassingly inefficient code. Similar issues
arise in the compilation of languages such as Haskell, where the
case-of-case transformation is employed to optimize dispatching on
algebraic datatypes~\citep{PEYTONJONES19983}.

Olivier Danvy, in his paper \emph{A New One-Pass Transformation into
Monadic Normal Form}~\citep{Danvy:2003fk}, presents an elegant family
of structurally recursive functions that translate Boolean expressions
and conditionals into a monadic normal form that (1) no longer
contains Boolean expressions, (2) the condition expression of a
condition is a value, and (3) jumps are expressed using zero-arity
function calls. One of the key parts of that translation is in the
compilation of a conditional expression such as $(\IF{b}{e_1}{e_2})$,
where the expression $b$ is compiled using a special function
$\mathcal{B}$ that takes two more arguments, the generated code for
the two branches, that is, for the two continuations of $b$. Thus,
$\mathcal{B}$ can analyze the expression $b$ to decide how to generate
code for the $\mathsf{if}$, or whether an $\mathsf{if}$ is even
necessary.
\[
\CE{\IF{b}{e_1}{e_2}} = \CB{b}{\CE{e_1}}{\CE{e_2}}
\]
For example, in the simple cases where $b$ is $\TRUE$ or $\FALSE$,
no branching is needed.
\begin{align*}
\CB{\TRUE}{c_1}{c_2} &= c_1 \\
\CB{\FALSE}{c_1}{c_2} &= c_2
\end{align*}

@@

\subsection{Structured Imperative IR}
\label{sec:imp}

@@

CFG = List CStmt

data CProg : Set where
  Program : $ℕ$ $→$ Id $→$ CFG $→$ CProg
\end{lstlisting}

Note that the \lstinline{CStmt} type represents not just one statement
but a sequence of statements that always end in either a
\lstinline{Return} statement (that halts the program) or a jump to
another \lstinline{CStmt} via \lstinline{Goto} or \lstinline{IfEq}.
The control flow graph \lstinline{CFG} is a list of \lstinline{CStmt},
and jump labels are indexes into that list.  So \lstinline{Program}
consists of (1) the number of variables, (2) the index of the start
node, (3) the control flow graph.

We choose a big-step semantics for C-like language because the
semantics is no longer structurally recursive and furthermore,
programs are no longer guaranteed to terminate.\footnote{The programs
in the image of Explicate Control are guaranteed to terminate, but
Agda doesn't know that.}  So encoding the semantics as a function in
Agda would come with complications (such as using a gas parameter)
that we side-step by using a big-step semantics. Compared to the
semantics for the imperative IR, this semantics includes a parameter
for the control flow graph.

\begin{lstlisting}
data _,_$⊢ᶜ$_$⇓$_$⊣$_ : StateImp $→$ CFG $→$ CStmt
                      $→$ Value $→$ StateImp $→$ Set
\end{lstlisting}

The rule for \lstinline{Goto} looks up the CFG node for label $l$ and
executes the sequence of statements in that node.
\[
\inferrule{s , G ⊢ᶜ G[l] ⇓ v ⊣ s′}
          {s , G ⊢ᶜ \mathtt{Goto}~ l ⇓ v ⊣ s′}
\]
\noindent The rule for \lstinline{IfEq}, in the case when the
arguments are equal, is as follows:
\[
\inferrule{\texttt{interp-CExp}(\mathtt{Eq}~a_1~a_2) = \mathsf{just}~\mathsf{true} \\
           s , G ⊢ᶜ G[l_1] ⇓ v ⊣ s′}
          {s , G ⊢ᶜ \mathtt{IfEq}~a_1~a_2~l_1~l_2 ⇓ v ⊣ s′}
\]

@@

\subsection{Remove Complex Operands is Correct}

@@

\subsection{Lift Locals is Correct}

@@

The Explicate Control pass is responsible for translating from the
structured control-flow of an abstract syntax \emph{tree} into a
language with a goto-based control flow \emph{graph}.\footnote{Our
source language does not include loops, so the graph happens to be a
directed acyclic graph.} In particular, Explicate Control translates
\lstinline{If} expressions, whose conditions may be arbitrary
expressions that include other \lstinline{If} expressions, into the
\lstinline{IfEq} form that only branches based on an equality test. In
so doing, Explicate Control brings together uses of the \lstinline{Eq}
comparison operator and the \lstinline{If} expressions that depend on
them, thereby avoiding the needless extra code that would be generated
by a naive approach.

We implement the Explicate Control pass with four functions in Agda,
the \lstinline{explicate} function translates a whole program from the
imperative IR to the C-like language, but it delegates all the work to
the other three functions. These three functions are organized
according to the context in which they are called. Reviewing the
grammar of \lstinline{Imp-Exp} (section~\ref{sec:imp}), there are
three contexts in which an \lstinline{Imp-Exp} may appear:
\begin{enumerate}
\item Tail position: the result of the expression is the output of the program,
\item Assignment: the result of the expression is assigned to a variable, and
\item Predicate: the (Boolean) result of the expression determines which
    branch to take in a conditional.
\end{enumerate}

The functions for Assignment and Predicate have extra parameters for
their continuations, one for Assignment and two for Predicate. The
continuations are represented simply as the generated code in the
C-like IL. The function for Tail position of course needs no
continuation parameter.

All three functions will be building a control flow graph, which needs
to be threaded into and out of these functions, so they use the
following a \lstinline{Build} monad:

\begin{lstlisting}
Build A = CFG $→$ A × CFG
retB : {A : Set} $→$ A $→$ Build A
_thenB_ : {A B : Set} $→$ Build A $→$ (A $→$ Build B)
           $→$ Build B
add-node : CStmt $→$ Build Id
\end{lstlisting}

\noindent Here are the signatures of the four functions that make up
the Explicate Control pass. Next we discuss select cases of these
functions.

\begin{lstlisting}
explicate : Imp-Prog $→$ CProg
explicate-tail : Imp-Exp $→$ Build CStmt
explicate-assign : Id $→$ Imp-Exp $→$ CStmt $→$ Build CStmt
explicate-pred : Imp-Exp $→$ CStmt $→$ CStmt $→$ Build CStmt
\end{lstlisting}

\paragraph{Explicate Tail}

Starting with \lstinline{explicate-tail}, in the case of an atomic
expression or primitive operator, we generate a return statement.
Here is the case for subtraction.

\begin{lstlisting}
explicate-tail (Sub a$₁$ a$₂$) = retB (Return (Sub a$₁$ a$₂$))
\end{lstlisting}

\noindent Moving on to the assignment statement, we see a pattern of
compiling the program in reverse. We first compile the continuation
\lstinline{e$₂$}, which is in tail position, producing the code \lstinline{c$₂$}.
We then compile \lstinline{e$₁$}, which is in assignment position,
passing \lstinline{c$₂$} as its continuation.

\begin{lstlisting}
explicate-tail (Assign x e$₁$ e$₂$) =
    (explicate-tail e$₂$) thenB
    $λ$ c$₂$ $→$ explicate-assign x e$₁$ c$₂$
\end{lstlisting}

\noindent Regarding a conditional expression, we first compile the two
branches, \lstinline{e$₂$} and \lstinline{e$₃$}, which are both in tail
position, and then compile the condition \lstinline{e$₁$}, which is in
predicate position, so we pass \lstinline{c$₂$} and \lstinline{c$₃$} in
the continuation parameters.

\begin{lstlisting}
explicate-tail (If e$₁$ e$₂$ e$₃$) =
  (explicate-tail e$₂$) thenB
  $λ$ c$₂$ $→$ (explicate-tail e$₃$) thenB
  $λ$ c$₃$ $→$ explicate-pred e$₁$ c$₂$ c$₃$
\end{lstlisting}

\paragraph{Explicate Assignments}

Turning to \lstinline{explicate-assign}, in each base case (for
atomics and primitive operators), we simply generate an assignment
statement with continuation \lstinline{c}.

\begin{lstlisting}
explicate-assign x (Atom a) c =
  retB (Assign x (Atom a) c)
\end{lstlisting}

The case for conditional expressions is interesting: we do not
directly generate an assignment statement. The two branches,
\lstinline{e$₂$} and \lstinline{e$₃$}, are in assignment position,
so we recursively compile them with \lstinline{explicate-assign}.
They both have the same continuation \lstinline{c} and the first rule
of compilation is that you cannot duplicate code.\footnote{GHC
violated this rule a decade ago:
\url{https://stackoverflow.com/questions/35815503/what-ghc-optimization-is-responsible-for-duplicating-case-expressions}.}
Thus, we add \lstinline{c} to the control-flow graph and
instead use a \lstinline{Goto} to its label as the continuation
for \lstinline{e$₂$} and \lstinline{e$₃$}.

\begin{lstlisting}
explicate-assign y (If e$₁$ e$₂$ e$₃$) c =
   add-node c thenB
   $λ$ l $→$ explicate-assign y e$₂$ (Goto l) thenB
   $λ$ c$₂$ $→$ explicate-assign y e$₃$ (Goto l) thenB
   $λ$ c$₃$ $→$ explicate-pred e$₁$ c$₂$ c$₃$
\end{lstlisting}

\paragraph{Explicate Predicates}

The base cases for \lstinline{explicate-pred} are interesting.  If the
expression is a literal \lstinline{true} or \lstinline{false}, then we
return the continuation for the \lstinline{thn} or \lstinline{els}
branch, respectively. (This may case \lstinline{thn} or
\lstinline{els} to become dead code, which we discuss later.)

\begin{lstlisting}
explicate-pred (Atom (Bool true)) thn els = retB thn
explicate-pred (Atom (Bool false)) thn els = retB els
\end{lstlisting}

\noindent In the case of a variable, we compare it to \lstinline{true}
and branch to either the \lstinline{thn} or \lstinline{els}
continuations. However, the syntax of \lstinline{IfEq} requires labels
to jump to, so we add both continuations to the CFG.

\begin{lstlisting}
explicate-pred (Atom (Var x)) thn els =
  add-node thn thenB
  $λ$ l$₁$ $→$ add-node els thenB
  $λ$ l$₂$ $→$ retB (IfEq (Var x) (Bool true) l$₁$ l$₂$)
\end{lstlisting}

\noindent The case for an equality comparison delivers on one of the
goals of Explicate Control, which is to combine comparison expressions
with branching whenever possible.

\begin{lstlisting}
explicate-pred (Eq a$₁$ a$₂$) thn els =
  add-node thn thenB
  $λ$ l$₁$ $→$ add-node els thenB
  $λ$ l$₂$ $→$ retB (IfEq a$₁$ a$₂$ l$₁$ l$₂$)
\end{lstlisting}

For the finale of the Explicate Control pass, we discuss the
translation of conditionals in \lstinline{explicate-pred}.  The two
branches \lstinline{e$₂$} and \lstinline{e$₃$} are in predicate
position, so we compile them with recursive calls to
\lstinline{explicate-pred} to obtain \lstinline{c$₂$} and
\lstinline{c$₃$}. However, they share the same two continuations,
\lstinline{thn} and \lstinline{els}, so we add both of them to the CFG
to avoid code duplication. Finally, we compile \lstinline{e$₁$} with
\lstinline{explicate-pred}, passing \lstinline{c$₂$} and
\lstinline{c$₃$} as the continuations.

\begin{lstlisting}
explicate-pred (If e$₁$ e$₂$ e$₃$) thn els =
   add-node thn thenB
   $λ$ l$₁$ $→$ add-node els thenB
   $λ$ l$₂$ $→$ explicate-pred e$₂$ (Goto l$₁$) (Goto l$₂$) thenB
   $λ$ c$₂$ $→$ (explicate-pred e$₃$ (Goto l$₁$) (Goto l$₂$)) thenB
   $λ$ c$₃$ $→$ explicate-pred e$₁$ c$₂$ c$₃$
\end{lstlisting}

\noindent The above handling of \lstinline{If} happens to perform the
case-of-case optimization~\citep{PEYTONJONES19983}.

\paragraph{Discussion}

Throughout the Explicate Control pass we added code to the
control-flow graph whenever we needed to avoid code
duplication. However, some further care is needed to avoid adding (1)
trivial nodes that only contain a \lstinline{Goto} and (2) nodes that
are never the target of a jump (i.e., dead code). The
\lstinline{add-node} function can easily handle (1) by inspecting its
argument and if it is a \lstinline{Goto}, just return its label
instead of adding a new node to the CFG.  Avoiding dead code (2) is
more challenging, either requiring a post-processing of the CFG or use
lazy evaluation to delay the addition of nodes to the CFG to the point
where they are definitely needed~\citep{Siek:2023tr}.

\subsection{Explicate Control is Correct}

The correctness theorem for Explicate Control says that if a program
output is $v$ given input $s$, then Explicate Control produces a
program that outputs the same value given the same input.

\begin{lstlisting}
explicate-correct : $∀$ {p}{s}{v}
  $→$ interp-imp p s v
  $→$ interp-prog (explicate p) s v
\end{lstlisting}

The correctness theorem delegates to three lemmas, one for each
\lstinline{explicate-*} function.

\paragraph{Explicate Tail is Correct}

The statement of the lemma for \lstinline{explicate-tail} is the most
straightforward. If expression \lstinline{e} compiles to
\lstinline{c}, and if \lstinline{e} evaluates to \lstinline{v}, then
\lstinline{c} also evaluates to \lstinline{v}:
\begin{lstlisting}
explicate-tail-correct :
  $∀${e}{$ρ$ $ρ'$}{s s'}{G G$′$}{c}{v}
  $→$ (s , $ρ$) $⊢$ e $⇓$ v $⊣$ (s' , $ρ'$)
  $→$ explicate-tail e G $≡$ (c , G$′$)
  $→$ (s , $ρ$) , G$′$ $⊢ᶜ$ c $⇓$ v $⊣$ (s' , $ρ'$)
\end{lstlisting}
\noindent The cases in the proof of \lstinline{explicate-tail-correct}
for assignment and conditional expressions rely on the correctness
lemmas for \lstinline{explicate-assign} and
\lstinline{explicate-pred}.

The case for conditional expressions also requires a lemma about the
monotonicity of the control flow graph.  The induction hypothesis for
\lstinline{e$₂$} says that the generated \lstinline{c$₂$} evaluates
using the CFG that was produced by the compilation of
\lstinline{e$₂$}, but to satisfy the induction hypothesis
for \lstinline{e$₁$} we need \lstinline{c$₂$} evaluated using the CFG
that was produced by the compilation of \lstinline{e$₃$}.
We define the following notation for
expression that one CFG is a prefix of another CFG:
\begin{lstlisting}
G$₁$ $↝$ G$₂$ = $∃$[ G ] G$₁$ ++ G $≡$ G$₂$
\end{lstlisting}
\noindent We prove the following lemma which states that the ouput CFG of
\lstinline{explicate-tail} is an extension of its input CFG.
\begin{lstlisting}
explicate-tail-CFG : $∀${e}{G$₁$ G$₂$}{c}
  $→$ explicate-tail e G$₁$ $≡$ (c , G$₂$)
  $→$ G$₁$ $↝$ G$₂$
\end{lstlisting}

\paragraph{Explicate Assign is Correct}

The correctness lemma for \lstinline{explicate-assign} adds another
premise which states that the continuation \lstinline{c$₂$} evaluates
in an environment that was updated by the assignment.

\begin{lstlisting}
explicate-assign-correct :
  $∀${x}{e$₁$}{c$₂$}{$ρ$ $ρ′$ $ρ″$}{s s$′$ s$″$}{G G$′$}{c$₁$}{n v}
  $→$ (s , $ρ$) $⊢$ e$₁$ $⇓$ n $⊣$ (s$′$ , $ρ′$)
  $→$ (s$′$ , update $ρ′$ x n) , G $⊢ᶜ$ c$₂$ $⇓$ v $⊣$ (s$″$ , $ρ″$)
  $→$ explicate-assign x e$₁$ c$₂$ G $≡$ (c$₁$ , G$′$)
  $→$ (s , $ρ$) , G$′$ $⊢ᶜ$ c$₁$ $⇓$ v $⊣$ (s$″$ , $ρ″$)
\end{lstlisting}

\noindent The case for conditional expressions needs a couple more
lemmas regarding control flow graphs. Similar to
\lstinline{explicate-tail}, we need a lemma which states that the
ouput CFG of \lstinline{explicate-assign} is an extension of its input
CFG.

\begin{lstlisting}
explicate-assign-CFG : ∀{x}{m}{t t'}{G₁ G₂}
  → explicate-assign x m t G₁ ≡ (t' , G₂)
  → G₁ ↝ G₂
\end{lstlisting}

\paragraph{Explicate Predicate is Correct}

...

\begin{lstlisting}
explicate-pred-correct :
  $∀$ {e$₁$}{c$₁$ c$₂$ c}{b}{v}{s s$′$ s$″$}{$ρ$ $ρ′$ $ρ″$}{G$′$ G$″$ G$‴$}
  $→$ (s , $ρ$) $⊢$ e$₁$ $⇓$ Bool b $⊣$ (s$′$ , $ρ′$)
  $→$ ((s$′$ , $ρ′$) , G$′$ $⊢ᶜ$ (if b then c$₂$ else c$₃$) $⇓$ v $⊣$ (s$″$ , $ρ″$))
  $→$ G$′$ $↝$ G$″$
  $→$ explicate-pred e$₁$ c$₂$ c$₃$ G$″$ $≡$ (c$₁$ , G$‴$)
  $→$ (s , $ρ$) , G$‴$ $⊢ᶜ$ c$₁$ $⇓$ v $⊣$ (s$″$ , $ρ″$)
\end{lstlisting}
